---
output:
  pdf_document: default
  html_document: default
---


# Assignment 0

```{r setup, include = FALSE}
setwd("C:/Users/joaom/Desktop/IO/")

rm(list=ls())

library(tidyverse) #for data manipulation
library(haven) #for reading .dta

```

Let's start importing data, naming variables and creating new ones where necessary.

```{r dataimport}
data <- read_dta("NEW7080.dta") %>%
  select(age = v1,
         ageq = v2,
         educ = v4,
         enocent = v5,
         esocent = v6,
         lwklywge = v9,
         married = v10,
         midatl = v11,
         mt = v12,
         neweng = v13,
         census = v16,
         qob = v18,
         race = v19,
         smsa = v20,
         soatl = v21,
         wnocent = v24,
         wsocent = v25,
         yob = v27) %>%
  mutate(cohort = if_else(40<=yob & yob < 50, "40-49",
                  if_else(30<=yob & yob < 40, "30-39",
                          "20-29")),
         ageqsq = ageq^2,
         cons = 1) %>%
  filter(cohort == "20-29")


for (quarter in (1:4)) {
  for(year in (1920:1929)){
    data[str_c("qt",quarter,"yr",year)] <- if_else(data$yob == year & data$qob == quarter, 1, 0)
    
  }
}

for(year in (1920:1929)){
  data[str_c("yr",year)] <- if_else(data$yob == year, 1, 0)
}
```

Here I define a function for OLS. It converts the inputs into matrices so we can pass dataframes directly.

```{r ols}
OLS <- function(X,Y){
  
  X <- as.matrix(X)
  Y <- as.matrix(Y)
  
  n = length(Y)
  k = dim(X)[2]

  beta <- solve(t(X) %*% X) %*% t(X) %*% Y
  
  fitted <- X %*% beta
  residual <- Y - fitted
  sigma <- sum(residual^2) / (n - k)
  var <- sigma * solve(t(X) %*% X)
  se <- sqrt(diag(var))
  
  results <- data.frame(beta = c(beta), se = se)
  
  return(results)
}
```

Here I do the same for IV.

```{r iv}
IV <- function(X,Z,Y){
  
  X <- as.matrix(X)
  Y <- as.matrix(Y)
  Z <- as.matrix(Z)
  
  n = length(Y)
  k = dim(X)[2]
  
  beta_fs <- solve(t(Z) %*% Z) %*% t(Z) %*% X
  
  X_dot <- Z %*% beta_fs
  
  beta <- solve(t(X_dot) %*% X) %*% t(X_dot) %*% Y
  
  fitted <- X %*% beta
  residual <- Y - fitted
  sigma <- sum(residual^2) / (n - k)
  var <- sigma * solve(t(X_dot) %*% X_dot) 
  se <- sqrt(diag(var))
  
  results <- data.frame(beta = c(beta), se = se)
  return(results)
}
```

Finally, select some data and print the results. First for OLS.

```{r olsresults}
Y <- data %>% select(lwklywge)

Xlist <- list(data %>% select(educ, yr1920:yr1928, cons),
              data %>% select(educ, ageq, ageqsq, yr1920:yr1928, cons),
              data %>% select(educ, race, married, smsa, neweng, midatl,enocent, 
                              wnocent, soatl,esocent, wsocent, mt, yr1920:yr1928, cons),
              data %>% select(educ, ageq, ageqsq, race, married, smsa, 
                              neweng, midatl, enocent, wnocent, soatl,
                              esocent, wsocent, mt, yr1920:yr1928, cons)
              )

tableIVa <- map(Xlist, OLS, Y)

print(tableIVa)
```

Now for IV. Note that the X variables are the same as before. For the second and fourth columns, we omit some variables due to collinearity.

```{r ivresults}
Zlist <- list(data %>% select(qt1yr1920:qt4yr1929),
              data %>% select(ageqsq, qt1yr1920:qt4yr1927, cons),
              data %>% select(race, married, smsa, neweng, midatl, enocent, wnocent,
                              soatl,esocent, wsocent, mt, qt1yr1920:qt4yr1928, cons),
              data %>% select(ageqsq, race, married, smsa, 
                              neweng, midatl, enocent, wnocent, soatl,
                              esocent, wsocent, mt, qt1yr1920:qt4yr1927, cons)
              )

tableIVb <- map2(Xlist, Zlist, IV, Y)

print(tableIVb)

```
In both cases, the results are the same as the ones obtained in Stata.

Now lets do the logit, using Newton's algorithm.

```{r Logit}
Logit <- function(X,Y, precision = 1e-8, maxiter = 1000){

  X <- as.matrix(X)
  Y <- as.matrix(Y)
  
  beta_new <- solve(t(X) %*% X) %*% t(X) %*% Y #start with ols as a guess
  
  p <- function(z) {1/(1+exp(-z))}

  iter <- 0
  while (iter < maxiter) {
    beta_old <- beta_new
    
    X_tilde <- X* (p(X%*%beta_old) %*% matrix(1,1,dim(X)[2]))
    
    beta_new <- beta_old + solve(t(X)%*%X_tilde) %*% t(X) %*% (Y - p(X%*%beta_old))
    iter <- iter + 1
    if (sum(abs(beta_new - beta_old)) < precision) {
      break
    }
  }
  
  return(beta_new)
}
```

And here's a function to get the partial effects of the first regressor. Since the usual procedure depends on whether or not it is a binary variable, the function can compute it either way.

```{r APE}
getPartialEffect <- function(beta, X, binary = FALSE) {
  P <- function(x){exp(x)/(1+exp(x))}
  if(binary == TRUE){
    X1 <- X %>% mutate(one = 1) %>% select(one, 2:length(X)) %>% as.matrix()
    P1 <- P(X1%*%beta)
    X0 <- X %>% mutate(zero = 0) %>% select(zero, 2:length(X)) %>% as.matrix()
    P0 <- P(X0%*%beta)
    APE <- mean(P1 - P0)
    return(APE)
  }
  if(binary == FALSE){
    X <- X  %>% as.matrix()
    APE <- mean(beta[1]*P(X%*%beta))
    return(APE)
  } 
}
```

Get data and print results.

```{r logitresults}
Y <- data %>% mutate(educ12 = if_else(educ >= 12,1,0)) %>% select(educ12)
X <- data %>% mutate(qtr1 = if_else(qob == 1, 1, 0)) %>% select(qtr1, ageq, ageqsq, cons)

beta <- Logit(X,Y)
print(beta)

APE <- getPartialEffect(beta,X,binary = T)
print(APE)

```

